\plaintitle

### 2016-05-21 paper/paper.tex
Reworked the paper, moved some stuff around, related work should be rewritten, then I think we're pretty much done

\documentclass{sigchi}
% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

@@

\begin{abstract}
    In this paper we describe an interactive visualization of the NBA competition over the last 30 years. With three different views, we give the opportunity for the user to explore various kinds of data about NBA seasons including standings, team statistics, transfers and player statistics. The d3.js framework was used to develop this visualization.
    %TODO add results

\end{abstract}

%TODO keywords
\category{H.5.m.}{Information Interfaces and Presentation
  (e.g. HCI)}{Miscellaneous} \category{See
  \url{http://acm.org/about/class/1998/} for the full list of ACM
  classifiers. This section is required.}{}{}

@@

The National Basketball Association (NBA) is the most famous and perhaps the
best professional basketball league in the world. Like in every other sports
branch, a lot of data and statistics about players, teams, games and seasons
are available, mainly in large tables. Despite these tables contains a lot of
information, it is very hard if not impossible, to interpret, compare
efficiently and gather insights in it. In our work we try to present this data
in a whole different way. Therefore we have focused ourselves on some parts and
not the entire available data.

This paper describes the visualization of NBA data made by the ``The Tufters''
team for the course ``Information Visualization''. In section~\ref{sec:goal} we
describe the goal of the visualization and the target audience. In
section~\ref{sec:data} we describe the data used, including its origins,
advantages and limitations. In section~\ref{sec:literature} we give an overview
of related literature and web resources, including related visualizations.
This includes both visualizations of NBA or other sports data, as well as
visualizations tackling an isue encountered during the development of our
visualization. In section~\ref{sec:visualization} we describe the visualization
itself. We give an overview of the different stages of development of the
visualization, as well as the major design decisions.
Section~\ref{sec:discussion} discusses potential improvements of the final
visualization and lessons learned from the project. We conclude in
section~\ref{sec:conclusion}. Appendix~\ref{sec:terminology} contains
definitions for common basketball terms used in this paper.

\section{Goal and Target Audience}\label{sec:goal} 
The visualization's goal can be best summarized by the following sentence:
``Why did a certain NBA-team win the Championship?'' The visualization focuses
on exposing relations between team performance and their player roster over
several years. It allows users to find explanations for major improvements and
declines in team performance.  The visualization allows exploration of NBA data
by lay persons.  More specifically, the visualization does not focus on premade
explanations for phenomenons visible in the NBA data. By providing easy and
intuitive access to the data, the visualization allows users to draw their own
conclusions. The visualization's target audience are lay persons. Specifically,
fans of NBA are the core of our target audience. This means the visualization
assumes most common basketball terms are known to the audience and does not
provide basic terminology explanations.

< We also give basic insights in why a
statistic changes (arrow views) we try to help them explore patters and trigger
them for questions Do we create awareness?  >

\section{Data}\label{sec:data}
The data visualized is a subset of the data available on basketball statistics
site basketball-reference\cite{basketball-reference}. A wide range of data is
available on this site. This includes common basketball performance statistics,
such as field goals, percentage of shots scored, shooting distance, minutes
played per season, number of personal fouls made, salaries and much more. These
statistics are avalable for each player individually for each season. Aggregated
data for entire teams and for entire seasons is also available. Additionally,
individual game statistics are also available, including playoff games.  In our
visualization we only use data from 1984 onwards. This is done because of rule
changes in the NBA which changed the number of teams competing and the
competition's structure. To simplify implementation of our visualization, only
data after the last major rule change was used.  The data we use includes league
standings and playoff rankings for each team, team overall statistics, the
team's roster and individual player statistics for each year, including the
PER~(Player Efficiency Rating)\cite{per}.

The data was gathered by scraping the basketball-reference site using the
provided download capabilities. Most of the data was downloaded in csv format,
while some tables had to be manually scraped. This process was automated using
Python scripts. The data was then combined in a preprocessing step. In this
step, each team's playoff rankings were calculated based on the matches played
during the playoffs, and the rest of the data was combined into json format. The
final preprocessing step combined all data into one json file.

<Tell something more about the data in terms of tall/wide data
Mention that it's a time series
Tell something more about pre-processing steps (why?, did we change it over time?)
rfr to picture in 1st ppt sl 60>

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/data}
  \caption{An overview of the data.}~\label{fig:data}
\end{figure}

\section{Related Work}\label{sec:literature}

@@

@@

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/playoffviewwithcontext}
  \caption{The play-off view without a team selected.}~\label{fig:playoffviewnoteam}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/playoffviewteamselected}
  \caption{The play-off view with a team selected.}~\label{fig:playoffviewteam}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/statisticsview}
  \caption{The statistics view.}~\label{fig:statisticsview}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/teamview}
  \caption{The team view.}~\label{fig:teamview}
\end{figure}

<Explain that this complies with the infovis seeking mantra: Overview, zoom\&filter, details on demand
play-off view = overview
statistics view= details on demand
Talk about th organization of the 3 views + the choice for a tiled approach
- Each page is a 'tile' that we force with fixed scrolling
- Use of the keyboard is possible (space bar to scroll + arrows, arrows for timeline)
- This papers are usefulle to refer to
http://courses.ischool.berkeley.edu/i247/f05/readings/Baldonado\_MultipleViews\_AVI00.pdf
http://www.interactiondesign.us/courses/2011\_AD690/PDFs/Shneiderman\_1996.pdf
http://www2.parc.com/istl/groups/uir/publications/items/UIR-1986-02-Mackinlay-TOG-Automating.pdf
>

All three are discussed in more detail below. The three views are available to
the user on a single webpage. The user start on the play-off view and is able to
navigate to the statistics and team views by scrolling down or clicking on
teams. A different season can be selected with the left and right arrows or by clicking on the timeline. The timeline is a fixed part of the visualization, hence visible in the three different views.
%The webpage is built in such a way that it gives a hint to the user at the bottom
%or the top of a page when another view is available. By showing a different 
%background color and an arrow, we invite the visitor to explore other views. The 
%user is also at liberty to scroll through these views. However, a fixed scroll is
%implemeted so users will only end in complete views when moving around. Note that 
%the user also will be guided through all 3 views when performing actions described 
%in the section below.

\subsection{The play-off view}
This view informs the user of the roster and end positions in the NBA play-offs for a selected season, in general or for a selected team. The view consists of two parts: 
\begin{itemize}
    \item Context section: the top part informs the user about the selected team and shows a legend of the play-off view.
    \item Bubble section: this section represents the play-offs of the selected season
\end{itemize}
The play-off view is shown in Figure~\ref{fig:playoffviewwithcontext}.

\subsubsection{Context section}
The context section contains a legend that explains the stroke colors used in
the bubble section. The stroke is gold, silver or bronze if the team ended
respectively first, second or third in the regular competition. Teams that
didn't end in a podium place during the regular competition have a red or blue
stroke color indicating the team's region, respectively the western or eastern
conference. The context section also contains a link to another page
(``About'') where the user can get an explanation of the visualization with its
different views, can get some background information about the NBA and the
terminolgy used, and can check the source of our used data. If the user hovers
over or selects a team in the bubble section, the context section shows more
information about that team, as shown in
Figure~\ref{fig:playoffviewteamselected}. This includes the Simple Rating
System score (SRS), the end position in the play-offs and the end position in
the regular competition.

\subsubsection{Bubble section}
This section presents the roster and end positions of the play-offs for a
selected season. Every team is respresented by a circle with the logo of the
team inside it. Our initial desing filled the circle with the logo of the team
and colored the circle red or blue depending on the region where the team
competed. The shade of colour was darker if their final ranking was higher in
their region. According to the Ranking of Perceptual
Tasks~\cite{perceptualranking} this is a good property for ordered values like
the regular competition ranking. But the combination of the logo's of the teams
with the colors made it to heavy and less easy to compare the end ranking of
different teams. As an alternative we tried the same view without the logo's,
but this is inconvenient because you don't know which team is represented by
which circle. Therefore we've chosen to only use the logo's of the teams as we
thought this is more user friendly and the extra information of the ranking
doesn't outweight. The size of the circle represents the SRS score using an own
interpretation of perceptual scaling because the (Flannery) Appearance
Compensation is not usable with a domain of both negative and positive values.
An exponential function with exponent 1.5 was used. This way the proportion
between the area of a bigger and the area of a smaller circle will be greater
than the proportion of values the areas represent. The stroke of the circle
illustrates the team's region or medal won during the regional competition as
explained in the context section. The curved lines between teams represent that
those teams have encountered each other in the playoffs. Each row represents an
end position in the play-offs. The higher a team is, the further it has come.
From the top to the bottom the rows represent the winner of the final (place
1), loser of the final (place 2), losers of the semi-finals (place 3), losers
of the quarter-finals (place 4), losers of the 1/8-finales (place 5). The last
row represents the teams that did not make it to the play-offs (place 6).  A
connection is always made between a team that is in a higher row and a team
that is in a lower row. The highest team of the two teams has won the
confrontation, and therefor went at least one round further in the play-offs.
The two up most teams are the winners of the play-offs in their own conference
and played the final. 
A more popular visualization of a play-off is the team double elimination
bracket.  Although this is a more common representation , and there are some
nice desings like in this NBA 2012 play-offs
visualisation~\cite{tournamentladder} where a sunburst visualisation is used,
we wanted to create a more compact version by eliminating the recurring
representation of a team. We considered enclosed circles, as used in this
visualisation of the wimbledon tournament roster~\cite{enclosedcircles}.
Although the enclosed circles is a compact representation as well, we have
continued with our own design as we believe this view is more original and
innovative. Furthermore we are convinced that our view gives a good
illustration of the end positions. It's generally assumed that the team that is
on top of the ``ladder'' is the winning team. We didn't encouter a similar view
when searching for alternatives to visualize a play-off or championship. A new
visualization has the advantage that it will trigger users as well.

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/presidentcandidatesvisualization}
  \caption{The path of president candidates in their road to become president.}~\label{fig:whitehousepath}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\columnwidth]{figures/playoffviewfirstdesign}
  \caption{First design of the play-off view.}~\label{fig:firstdesignplayoffview}
\end{figure}

@@

<Talk about the importance of estethics seen in the lesson>
Our initial play-off view was too heavy. It connected teams that played against
each other directly. When searching for alternatives to reflect on our choice we
discovered a similar visualization in a totally different context. 
<https://source.opennews.org/en-US/articles/nyts-512-paths-white-house/>
<comparison of original play-off view with white house paths>
This visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to clearer indicate how teams competed
to become NBA champion.
Next to the static view, the user can interact with the visualization. When he 
hoovers over a circle of a team, only the games played by that team are shown by highliting the 
competitors (their circles) and the curved lines that connect them. This view gets 
fixated when the users clicks on a team.
He/she then automatically gets forwarded to the statistics view where the play-off 
view is still represented in the right upper corner to make sure the user keeps
an overview. The statistics view will be explained in more detail in the next section.

< Hierarchy = sorting = facilitate trends of winners
position have a meaning in the play-off view
when selecting circle => zoom/more detail
Bubble principle of 
- Symmetry (whole = play-off)
- conectedness arc = games played
Pre attentive characteristics
- size of circles
- hue to highligt selections or indicate increase decline arrows next view

We considered a map view to just show the teams and there regular competition
ranking. But this was a view that has no additional value in comparison with
the play-off view except for the geographical position, which has no importance
for our visualisation. Furthermore we've found a visualisation of the NBA with
a map view~\cite{mapviewvisualization} similar to our idea, so we decided to
not implement this map view.

This visualization helped us discover data quality issues => cut off year 1984
>

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been
influenced in the selected season compared to the season before. The statistics
view is showed in Figure~\ref{fig:statisticsview}. The view consists of 3
coordinated parts:
\begin{itemize}
    \item Context section: the top part informs the user of the context of the selected team
    \item Arrow section: this section illustrates why a statistic changed compared to the previous season
    \item Small multiples section: small multiples of team statistics and how they evolve over time
\end{itemize}

\subsubsection{Context section}
The context section informs the user at all times of the current selection of a
team  that have been made in the play-off view. The statistics view will always
be based on a specific team and season that have been selected or that are
initated. This should be clear based on the textual information in the top
left. Next to that, in the top right a small play-off view is shown. It
indicates the team selected and how well it performed in  the play-offs during
that season, as explained in the play-off view section. When changing teams
and/or season, this view gets updated as well as the textual information.
<Explain that the context supports the short term memory of human beings rfr.
Monkey Illusion>

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team
compared with the season before. This should give a visitor more insight into
why a statistic changed over time. The selected team is represented by a circle
in the same way as explained in the play-off view. The change for a statistic
is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the influence of players who joined the team (team member current season, not previous season)
    \item Arrow on the right of the circle: indicating how the team inherent changed (team member this season and previous season)
    \item Arrow on the left pointing away from the circle: indicating the influence of players who left the team (team member previous season, not current season)
\end{itemize}
<figure of the statistics view>
The shirts next to the arrows are the shirts of players that had the biggest 
contribution to that arrow. Note that these shirts don't change in size. The box on 
the right enables the user to change between statistics. Next to that it informs the user of which statistic has been chosen. When the user hits the circle in the middle representing the teams, he is guided to the team view. This view will be explained in more detail in the next section.
<filtering: 
- time \& team
- based on exploration (small multiples)
- dynamically: changes impact multiple parts>

\subsubsection{Small multiples of different statistics}
Below the arrow section small multiples are shown of multiple statistics over time. 
The line charts enables a user to identify peaks or drops that possibly influenced the 
outcome of a team is this season, following seasons or previous seasons. A user interacts with the visualization by shifting the bar indicating the current year displayed. By doing this the view gets updated and the user can see the impact on the team in the context section and why the statistic changes in the arrow section. An alternative to scroll through time is to use the timeline at the bottom of the page.
<this vis helps us comparre statistics over time and relations between them
eg. srs score \& play-off ranking>

\subsection{The team view}
This view as well as the statistics view should give the user the ability to search
for explanations why a team is is performing better or worse over seasons. On the 
other hand, a user can see what impact a change in team characteristics has on its
overall performance.
This view shows each player of the team on its field positions. Shirts aligned next to each other will thus share the same position. The size of a shirt would illustrate a players performance on this spot. The PER score is initiated. This view should give
a user insight on how strong or weak a particular team is for a specific field position.
A user can then scroll through time, see how field positions evolve and what the influence is over time.
Initially we designed this view more elaborate. When the user would click on a shirt,
specific player information would be shown. Next to some general text information, 
we would have visualised a number of player statistics as small multiples. A bell curve should have informed how a player scores compared to his team, other players in the league on this position or all other players in the league. We considered other charts for the small multiples as well. A box plot or bullet was less clear for us to indicate where a player stands compared to the group he was compared with. Alternatives for the small multiples were evaluated as well. One good alternative was the shooting signature.\cite{peterbeshai}
However, creating such a visualization is again a project on itself and due to time 
constraints we left it aside. Other possibilities such as a heatmap or spider web didn't fulfill our need to give the visitor a clear overview of how a player performed compared to its peergroup.
<Sketch of the small multiples>

\subsection{Improving the result}
Talk about how we improved the data-ink ration in the end..

\section{Possible improvements}
As with most projects, the result is never 'finished', meaning that there are always improvements possible but due to timing or resource constraints the haven't all been implemented. In what  follows, we suggest a number of features that might improve the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience. For example when changing year, a transformation of the arrow in the statistics view would help the user see a difference between years.
    At the moment the arrows change at the blink of an eye, making it difficult for us humans with a short term memory to make the comparison.
    \item To further support our short term memory we could also 'save' the previous state. Eg. in the field view we could visualize ghost shirts to indicate which players left the team or position and highlight players that joined the team. 
    This would help the user see how the team changes over time.
    \item At the moment no attention has been payed to the process \& provenance of the exploration. A user thus needs to  memorize his action to be able to share or reproduce his actions. This could be improved by adding URL parameters. This
    way a user at least can share or save a state of an analysis. One step further is to provide a play-feature. The visitor could them share his exploration and replay it. A simple implemention would ask the visitor for a start date and an end date and
    would then 'play' the changes of a particular team and characteristic over time.
    \item We could support users further in exploring for patterns by focussing more on comparisons. We could integrate a view where a visitor is able to compare two teams in the same year, or in two different years.
\end{itemize}

\subsection{Technology}
To create the visualization, the d3 javascript framework~\cite{d3.js} was used in combination with html5~\cite{html5} and jquery~\cite{jquery}.  Note that D3 is based on HTML, CSS, Javascript and SVG.
Next to that,the use of D3 was a requirement for the project. This allows easy
access to the visualization as most modern browsers are capable of handling
these webstandards. 
The choice not to use the d3 framework for the entire visualization was made to
ease the layout configuration of the visualization. Instead, html was used to do
the global layout of the visualization. We opted to structure the site with multiple
divs as containers. To each container we allocate a svg with a specific visualization.
Eg. The statistic view is build out of x divs filled with different svg's created
with the d3 framework.
<sketch of how the page is structured>
When working with d3.js we encountered some obstacles:
- A synchrounous call need to be combined with synchronous calls. More specific, when a visitor uses the timeline to scroll through time, the whiping of previous visualizations was not synced with the creation of the visualisation. This resulted in whiping sections when a visualization was not created yet. Hence the view resulted in multiple figures overlapping.

@@

@@

>>>>

\documentclass[chi_draft]{sigchi}
% Use this command to override the default ACM copyright statement
% (e.g. for preprints).  Consult the conference website for the
% camera-ready copyright statement.

@@

\begin{abstract}
    Visualization of sports data is very common in recent years as access to
    data has increased. The abundance of data and audience allows for new and
    innovative presentations. In this paper we describe an interactive
    visualization of the NBA competition over the last 30 years. With three
    different views, we allow the user to explore data about NBA seasons
    including standings, team statistics, transfers and player statistics. The
    d3.js framework was used to develop this visualization. 
\end{abstract}

%TODO keywords
\category{H.5.m.}{Information Interfaces and Presentation
  (e.g. HCI)}{Miscellaneous} \category{}{}{}

@@

The National Basketball Association (NBA) is the most famous professional
basketball league in the world\cite{nbawiki}. As in every other sports
branch, a lot of data and statistics about players, teams, games and seasons
are available, mainly in large tables. Despite the large amount of data
available, gathering useful insights from this data can be surprisingly
challenging. In our work we try to present this data in an innovative and
intuitive way, allowing the user to draw meaningful conclusions.

In section~\ref{sec:goal} we describe the goal of the visualization and the
target audience. In section~\ref{sec:data} we describe the data used, including
its origins, advantages and limitations. In section~\ref{sec:literature} we give
an overview of related literature and web resources, including related
visualizations.  This includes both visualizations of NBA or other sports data,
as well as visualizations tackling an isue encountered during the development of
our visualization. In section~\ref{sec:visualization} we describe the
visualization itself. We give an overview of the different stages of development
of the visualization, as well as the major design decisions.
Section~\ref{sec:discussion} discusses potential improvements of the final
visualization and lessons learned from the project. We conclude in
section~\ref{sec:conclusion}. Appendix~\ref{sec:terminology} contains
definitions for common basketball terms used in this paper.

\section{Goal and Target Audience}\label{sec:goal} 
The visualization's goal can be best summarized by the following sentence:
``Why did a certain NBA-team win the Championship?'' The visualization focuses
on exposing relations between team performance and their player roster over
several years. It allows users to find explanations for major improvements and
declines in team performance.  The visualization allows exploration of NBA data
by lay persons.  More specifically, the visualization does not focus on premade
explanations for phenomenons visible in the NBA data. By providing easy and
intuitive access to the data, the visualization allows users to draw their own
conclusions. The visualization's target audience are lay persons. Specifically,
basketball fans are the core of our target audience. It can be of particular
interest with the recent surge in fantasy leagues, for which data analysis is
paramount to a player's success~\cite{fantasy,fantasyskill}.

\section{Data}\label{sec:data}
The data visualized is a subset of the data available on basketball statistics
site basketball-reference\cite{basketball-reference}. A wide range of data is
available on this site. This includes common basketball performance statistics,
such as field goals, percentage of shots scored, shooting distance, minutes
played per season, number of personal fouls made, salaries and much more. All of
this data is available as time-series. These
statistics are avalable for each player individually for each season. Aggregated
data for entire teams and for entire seasons is also available. Additionally,
individual game statistics are also available, including playoff games.  In our
visualization we only use data from 1984 onwards. This is done because of rule
changes in the NBA which changed the number of teams competing and the
competition's structure. To simplify implementation of our visualization, only
data after the last major rule change was used.  The data we use includes league
standings and playoff rankings for each team, team overall statistics, the
team's roster and individual player statistics for each year, including the
PER~(Player Efficiency Rating)\cite{per}.

The data was gathered by scraping the basketball-reference site using the
provided download capabilities. Most of the data was downloaded in csv format,
while some tables had to be manually scraped. This process was automated using
Python scripts. The data was then combined in a preprocessing step. In this
step, each team's playoff rankings were calculated based on the matches played
during the playoffs, and the rest of the data was combined into json format. The
final preprocessing step combined all data into one json file. A model of the
available data can be seen on figure~\ref{fig:data}.

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/data}
  \caption{An overview of the data.}
  \label{fig:data}
\end{figure}

\todo{Rework realted work}
\section{Related Work}\label{sec:literature}

@@

(From bubble view)
A more traditional visualization of a play-off is the team double elimination
bracket.  Although this is a more common representation , and there are some
nice desings like in this NBA 2012 play-offs
visualisation~\cite{tournamentladder} where a sunburst visualisation is used,
we wanted to create a more compact version by eliminating the recurring
representation of a team. We considered enclosed circles, as used in this
visualisation of the wimbledon tournament roster~\cite{enclosedcircles}.
Although the enclosed circles is a compact representation as well, we have
continued with our own design as we believe this view is more original and
innovative. Furthermore we are convinced that our view gives a good
illustration of the end positions. It's generally assumed that the team that is
on top of the ``ladder'' is the winning team. We didn't encouter a similar view
when searching for alternatives to visualize a play-off or championship. A new
visualization has the advantage that it will trigger users as well.

(From field view)
One good alternative was the shooting signature.\cite{peterbeshai} However,
creating such a visualization is again a project on itself and due to time
constraints we left it aside. Other possibilities such as a heatmap or spider
web didn't fulfill our need to give the visitor a clear overview of how a player
performed compared to its peergroup.

@@

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/playoffviewwithcontext}
  \caption{The play-off view without a team selected.}~\label{fig:playoffviewnoteam}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/playoffviewteamselected}
  \caption{The play-off view with a team selected.}~\label{fig:playoffviewteam}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/statisticsview}
  \caption{The statistics view.}~\label{fig:statisticsview}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/teamview}
  \caption{The team view.}
  \label{fig:teamview}
\end{figure}

This technique follows the standards of information visualization. It provides
an overview, allows zooming and filtering and provides details on
demand\cite{mantra,multipleviews,automatingdesign}. The play-off view provides
the full overview which allows filtering by team. The statistics view provides
the ability to zoom into one specific team. There some details are provided
instantly, while others can be procured by hovering over certain elements.

The organization of these three views is in a vertical layout with fixed
scrolling. Navigation with the keyboard is supported as well.

The choice was made to use a thematic background instead of opting for the
traditional minimalistic approach to information visualization. This was done to
improve the attractiveness of the visualisation on the one hand and the users'
memory of it on the other hand\cite{aesthetics}.

All three views are discussed in more detail below. The three views are
available to the user on a single webpage. The user start on the play-off view
and is able to navigate to the statistics and team views by scrolling down or
clicking on teams. A different season can be selected with the left and right
arrows or by clicking on the timeline. The timeline is a fixed part of the
visualization. It is visible in the three different views.

\subsection{The play-off view}
This view informs the user of the roster and end positions in the NBA play-offs
for a selected season, in general or for a selected team. The view consists of
two parts: 
\begin{itemize}
    \item Context section: the top part informs the user about the selected team
    and shows a legend of the play-off view.
    \item Bubble section: this section represents the play-offs of the selected
        season.
\end{itemize}
The play-off view is shown in figure~\ref{fig:playoffviewwithcontext}.

\subsubsection{Context section}
The context section contains a legend for the bubble section. The bubble's
borders are gold, silver or bronze if the team ended respectively first, second
or third in the regular competition. Teams that didn't end in a podium place
during the regular competition have a red or blue stroke color indicating the
team's region, respectively the western or eastern conference. The context
section also contains a link to an ``about'' page where the user can get an
explanation of the visualization with its different views, can get some
background information about the NBA and the terminolgy used, and can check the
source of our used data. If the user hovers over or selects a team in the bubble
section, the context section shows more information about that team, as shown in
Figure~\ref{fig:playoffviewteamselected}. The information shown includes the
Simple Rating System score (SRS), the end position in the play-offs and the end
position in the regular competition.

\subsubsection{Bubble section}
This section presents the roster and end positions of the play-offs for a
selected season. Every team is respresented by a circle with the logo of the
team inside it. The size of the circle represents the SRS score using an own
interpretation of perceptual scaling because the (Flannery) Appearance
Compensation is not usable with a domain of both negative and positive values.
An exponential function with exponent 1.5 was used. This way the proportion
between the area of a bigger and the area of a smaller circle will be greater
than the proportion of values the areas represent.

The curved lines between teams represent that those teams have encountered each
other in the playoffs. Each row represents an end position in the play-offs. The
higher a team is, the further it has come.  From the top to the bottom the rows
represent the winner of the final (rank 1), loser of the final (rank 2), losers
of the semi-finals (rank 3), losers of the quarter-finals (rank 4), losers of
the eighth finals (rank 5). The last row represents the teams that did not make
it to the play-offs (rank 6).

This simple presentation intuitively puts the winning teams higher than the
losing teams. The winner of a certain game can be easily deriving from their
relative vertical position. It is also clear that each team neads to have played
and won a series of matches against one other team at every level below
them~\ref{fig:playoffviewteamselected}. This view is further reinforced by
highlighting a team and its matches and opponents when hovering over it.
Additional team information is then also shown in the context part.

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/presidentcandidatesvisualization}
  \caption{The path of president candidates in their road to become president.}~\label{fig:whitehousepath}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/playoffviewfirstdesign}
  \caption{First design of the play-off view.}~\label{fig:firstdesignplayoffview}
\end{figure}

@@

\subsubsection{Previous designs}
Our initial design used the team circles' lightness to indicate their final
ranking in their region. The hue was used to show the league they were part of.
According to the Ranking of Perceptual Tasks~\cite{perceptualranking} this is a
good property for ordered values like the regular competition ranking. But the
combination of the logo's of the teams with the colors resulted in an overload
of informaiton and made it more difficult to compare the end ranking of
different teams.

As an alternative we tried the same view without the logo's. This was deemed
inconvenient because you cannot instantly see which team is represented by which
circle. Therefore we've chosen to only use the teams' logos.

Our initial play-off view was too dense because of the connections between teams
as well. It connected teams that played against each other directly. When
searching for alternatives to reflect on our choice we discovered a similar
visualization in a totally different context\cite{whitehousepath}.  This
visualization gives a cleaner view of the competitors for a specific team
compared to our original sketch. Hence we decided to adapt our visualization. We
added an intermediate step between two teams to more clearly indicate how teams
competed to become NBA champion.

\todo{Hierarchy = sorting = facilitate trends of winners
position have a meaning in the play-off view
when selecting circle => zoom/more detail
Bubble principle of 
- Symmetry (whole = play-off)
- conectedness arc = games played
Pre attentive characteristics
- size of circles
- hue to highligt selections or indicate increase decline arrows next view}

We initially also considered a map view to show the teams and their regular
competition ranking on a map of the USA. However, this view had no additional
value compared to the play-off view except for the geographical position, which
has no importance for our visualisation. Furthermore we've found a visualisation
of the NBA with a map view~\cite{mapviewvisualization} similar to our idea, so
we decided to not implement this map view.

\subsection{The statistics view}
The statistics view gives a user a clear overview of how a statistic has been
influenced in the selected season compared to the season before. The statistics
view is showed in Figure~\ref{fig:statisticsview}. The view consists of 3
coordinated parts:
\begin{itemize}
    \item Context section: the top part informs the user of the context of the
        selected team
    \item Arrow section: this section illustrates why a statistic changed
        compared to the previous season
    \item Selected statistics section: team statistics and how they evolve over
        time
\end{itemize}

\subsubsection{Context section}
The context section is partly identical to that of the play-offs view. The
team's information is also displayed in the top left. In the top right, the
bubble section of the play-offs view is displayed in a reduced, minimalistic
form. This view is repeated to create a sense of continuity between the
different views and provide context of the current position in the
visualization. The team shown in the statistics view is highlighted on this
smaller bubble section as well. Replicating this section also alleviates
short-term memory issues. It prevents the need to scroll back and forth between
the different views constantly.

\subsubsection{Arrow section of selected statistic}
This section illustrates a change for a chosen statistic for the selected team
compared to the previous season. This gives a user more insight into why a
statistic changed over time and how this change has impacted team performance.
The selected team is represented by a circle in the same way as explained in the
play-off view. The change for a statistic is represented by three arrows:
\begin{itemize}
    \item Arrow on the left pointing towards the circle: indicating the
        influence of players who joined the team 
    \item Arrow on the right of the circle: indicating how the team internally
        changed, e.g. by training its players.
    \item Arrow on the left pointing away from the circle: indicating the
        influence of players who left the team 
\end{itemize}
\todo{add statistics view image here}
The jerseys next to the arrows are the jerseys of players that had the biggest
contribution to that arrow. On the left, different statistics , including player
PER can be chosen from. Below this, player information is shown when hovering
over a player's jersey.

\subsubsection{Selected statistics}
Below the arrow section some selected statistics are shown over
time.  The line charts enables a user to identify peaks or drops that
influenced the outcome of a team is this season, following seasons or previous
seasons. A user interacts with the visualization by shifting the bar indicating
the current year displayed. By doing this the view gets updated and the user can
see the impact on the team in the context section and why the statistic changes
in the arrow section. This view provides a lot of additional information to
users by allowing them to see correlations between statistics and phenomena
quickly and easily.

\subsection{The team view}
This view shows each player of the team on its field positions. Players are
represented by their jerseys. Jerseys aligned next to each other share the same
position, which is indicated as well. The size of a jersey would illustrate a
players performance on this spot. This view as well as the statistics view
should give the user the ability to search for explanations why a team is is
performing better or worse over seasons. On the other hand, a user can see what
impact a change in team characteristics has on its overall performance. This
view should give a user insight on how strong or weak a particular team is for a
specific field position. A user can then scroll through time to see how field
positions evolve and what the influence is over time.

\subsubsection{Previous designs}
Initially we designed this view more elaborate. When the user would click on a
jersey, specific player information would be shown. A number of player
statistics would be shown as small multiples. A bell curve should have informed
how a player scores compared to his team, other players in the league on this
position or all other players in the league. We considered other charts for the
small multiples as well. A box plot or bullet was less clear for us to indicate
where a player stands compared to the group he was compared with. Alternatives
for the small multiples were evaluated as well. This part of the visualization
was deemed less important than the others and has not been fully developed
because of time and resource constraints. \todo{sketch of small multiples
here}

\subsection{Improving the result}\todo{Is this necessary? I don't know what to
write here}
Talk about how we improved the data-ink ration in the end..

\section{Future work}
As with most projects, the result is never 'finished', meaning that there are
always improvements possible but due to timing or resource constraints they
haven't all been implemented. We suggest a number of features that might improve
the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience.
        For example when changing year, a transformation of the arrow in the
        statistics view would help the user see a difference between years.  At
        the moment the arrows change at the blink of an eye, making it difficult
        for us humans with a short term memory to make the comparison. This
        would have been especially useful in the transitions between different
        years on the bubble view.
    \item To further support our short term memory we could also 'save' the
        previous state. Eg. in the field view we could visualize ghost jerseys
        to indicate which players left the team or position and highlight
        players that joined the team.  This would help the user see how the team
        changes over time.
    \item At the moment no attention has been payed to the process, provenance
        and history of the user's exploration session. A user thus needs to
        memorize his action to be able to share or reproduce his actions. This
        could be improved by adding URL parameters. This way a user at least can
        share or save a state of an analysis. One step further is to provide a
        play-feature. The visitor could them share his exploration and replay
        it. A simple implemention would ask the visitor for a start date and an
        end date and would then 'play' the changes of a particular team and
        characteristic over time.
    \item We could support users further in looking for patterns by focussing
        more on comparisons. We could integrate a view where a visitor is able
        to compare two teams in the same year, or in two different years.
\end{itemize}

\subsection{Technology}
To create the visualization, the d3 javascript framework~\cite{d3.js} was used
in combination with html5~\cite{html5} and jquery~\cite{jquery}.  Note that D3
is based on HTML, CSS, Javascript and SVG. This allows easy access to the visualization as
most modern browsers are capable of handling these webstandards.  The choice not
to use the d3 framework for the entire visualization was made to ease the layout
configuration of the visualization. Instead, html was used to do the global
layout of the visualization. We opted to structure the site with multiple divs
as containers. To each container we allocate a svg with a specific
visualization.

\todo{sketch of how the page is structured here}
When working with d3.js we encountered some obstacles:
\begin{itemize}
        \item A synchrounous call need to be combined with synchronous calls.
            More specific, when a visitor uses the timeline to scroll through
            time, the whiping of previous visualizations was not synced with the
            creation of the visualisation. This resulted in whiping sections
            when a visualization was not created yet. Hence the view resulted in
            multiple figures overlapping.
\end{itemize}

@@

\paragraph{NBA} National Basketball Association - The largest, most well-known
basketball of the USA.

@@

\paragraph{SRS} Simple Rating System - a rating that takes into account average
point differential and strength of schedule. The rating is denominated in points
above/below average, where zero is average\cite{srs}.

\paragraph{Playoffs} The National Basketball Association (NBA) playoffs are a
best-of-seven elimination tournament among 16 teams in the Eastern Conference
and Western Conference (called divisions, pre-1970), ultimately deciding the
winner of the NBA Finals\cite{playoffs}.

\paragraph{Jersey} The shirt worn by basketball players.
